TGI-spinal-power

### 2022-06-01 analysis/tgispine_zoib_power-sim.Rmd
adding zeros to simulated data set and correcting some bugs in the function

# finally, count how many zeros there are in the data
# this is to model zero inflated regression
# cold cnt
CNTc_zero <- df_VAS %>% 
    filter(VAS == 'cold' & manipulation == 'CNT') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# cold tgi
TGIc_zero <- df_VAS %>% 
    filter(VAS == 'cold', manipulation == 'TGI') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# warm cnt
CNTw_zero <- df_VAS %>% 
    filter(VAS == 'warm', manipulation == 'CNT') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# warm tgi
TGIw_zero <- df_VAS %>% 
    filter(VAS == 'warm', manipulation == 'TGI') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))

@@

@@

summary(mc.og)
summary(mw.og)
summary(mb.og)
```

```{r}
simulate_subj <- function(subid, llim, ulim) {
# trial info
trialN = 12
conditionN = 8
ratingN = 3
# creating simulation data-frame
df <- data.frame(matrix(ncol = 0, nrow = trialN*conditionN*ratingN))
df$id <- rep(subid,trialN*conditionN*ratingN)
df$trialn <- rep(1:trialN,conditionN*ratingN)
df$dermatome <- rep(rep(c(rep("within",trialN*2), rep("across",trialN*2)),2),ratingN)
df$cold_location <- rep(rep(c(rep("dist_rostr",trialN), rep("prox_caud", trialN)),conditionN/2),ratingN)
df$stimuli <- rep(c(rep("tgi",trialN*conditionN/2), rep("non-tgi",trialN*conditionN/2)),ratingN)
df$quality <- c(rep("cold",trialN*conditionN), rep("warm",trialN*conditionN), rep("burn",trialN*conditionN))

@@

# First, simulate one data-set and run models (warm, cold and burn perception) to take a look at data
```{r}
df_sim <- data.frame()
n = 50 #an example number of participants
for (i in 1:n) {
  df <- simulate_subj(i, 0, 1)
  df_sim <- rbind(df_sim,df)
}
df50 <- df_sim

# Plotting data distributions to see what kind of model family is required
df50$quality <- factor(df50$quality, 
                             levels = c("cold", "warm", "burn"))
ggplot2::ggplot(data = df50) +
  geom_density(aes(rating, fill = stimuli), alpha = .5) +
  facet_wrap(~quality) +
  theme_classic()
```

@@

# Look at model assumptions
```{r}
# cold assumption - on n = 50
mc.assump <- simulateResiduals(mc, n = 1000)
plot(mc.assump)
```

## SIMULATIONS ##
# Run through n = 1000 simulations of the pilot data, and extract p-value for the smallest effect (cold_location) of each model
# Extract the power from each model - first cold
```{r}
# the warm model has the smallest effect size, so should extract power calculation for this model only
# do this for 40 participants, for 1000 simulations
nSims <- 1000
n <- 40 # number of participants to simulate
po <- data.frame()
for (i in 1:nSims){
  # creating individual data-frames from simulated data
  df_sim <- data.frame()
  for (i in 1:n) {
    df <- simulate_subj(i, 0, 1)
    df_sim <- rbind(df_sim,df)
    }
  # then run the model based on simulated data
  m <- model(df_sim, 'cold')
  m1 <- summary(m)
  # extracting the p-value and effect size
  tmp <- c(m1$coefficients$cond[4,4], m1$coefficients$cond[4,1])
  po <- rbind(po,tmp)
}

# rename columns
names(po)[1] <- 'pval'
names(po)[2] <- 'eff'
po$sig <- as.numeric(po$pval < .05) #is significant
po <- po[!is.na(po$sig) ,] #removing any NAs (models that did not converge) from the data frame

power <- sum(po$sig)/length(po$sig) # calculating power

# some renaming
po_c <- po
power_c <- power

power_c
write.csv(po_c, 'simulation_model_cold.csv', row.names = FALSE)
```

# Then warm
```{r}
# the warm model has the smallest effect size, so should extract power calculation for this model only
# do this for 40 participants, for 1000 simulations
nSims <- 1000
n <- 40 # number of participants to simulate
po <- data.frame()
for (i in 1:nSims){
  # creating individual data-frames from simulated data
  df_sim <- data.frame()
  for (i in 1:n) {
    df <- simulate_subj(i, 0, 1)
    df_sim <- rbind(df_sim,df)
    }
  # then run the model based on simulated data
  m <- model(df_sim, 'warm')
  m1 <- summary(m)
  # extracting the p-value and effect size
  tmp <- c(m1$coefficients$cond[4,4], m1$coefficients$cond[4,1])
  po <- rbind(po,tmp)
}

# rename columns
names(po)[1] <- 'pval'
names(po)[2] <- 'eff'
po$sig <- as.numeric(po$pval < .05) #is significant
po <- po[!is.na(po$sig) ,] #removing any NAs (models that did not converge) from the data frame

power <- sum(po$sig)/nSims
# some renaming
po_w <- po
write.csv(po_w, 'simulation_model_warm.csv', row.names = FALSE)
power_w <- power

power_w
```

>>>>

# finally, count how many zeros there are in the data
# this is to model zero inflated regression
# cold cnt
CNTc_zero <- df_VAS %>% 
    filter(VAS == 'cold' & manipulation == 'CNT') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# cold tgi
TGIc_zero <- df_VAS %>% 
    filter(VAS == 'cold', manipulation == 'TGI') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# warm cnt
CNTw_zero <- df_VAS %>% 
    filter(VAS == 'warm', manipulation == 'CNT') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# warm tgi
TGIw_zero <- df_VAS %>% 
    filter(VAS == 'warm', manipulation == 'TGI') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# burn cnt
CNTb_zero <- df_VAS %>% 
    filter(VAS == 'burn', manipulation == 'CNT') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))
# burn tgi
TGIb_zero <- df_VAS %>% 
    filter(VAS == 'burn', manipulation == 'TGI') %>% 
    dplyr::count(beta == 0) %>% 
    mutate(prop = n / sum(n))

@@

# Proportion zero function
Replaces proportion of data with zero, based on proportions for each condition calculated from pilot data - used in the simulation
```{r}
set.seed(1)
prop_zero <- function(x, vas = c('cold','warm','burn')) {
  x2 <- data.frame()
  # randomly generate two numbers around a normal distribution of 0, with a standard deviation of .03
  r <- rnorm(2, mean=0, sd=.05)
  if (vas[1] == 'cold'){
    x1 <- x %>%  filter(quality == 'cold')
    # add this number to the proportion of zeros calculated for each quality & stimuli from pilot data, for each participant
    zp_cnt <- CNTc_zero$prop[2] + r[1] 
    zp_tgi <- TGIc_zero$prop[2] + r[2]
    # adding if loop to make sure zp does not reach 0
    if (isTRUE(zp_cnt < 0)){
      zp_cnt = 0
    }
    if (isTRUE(zp_tgi < 0)){
      zp_tgi = 0
    }
    # turn zp into the proportion of total columns
    zp_cnt <- round(zp_cnt*nrow(filter(x1, stimuli == 'non-tgi')))
    zp_tgi <- round(zp_tgi*nrow(filter(x1, stimuli == 'tgi')))

    # isolate TGI and non-TGI
    x_cnt <- x1 %>% filter(stimuli == 'non-tgi')
    x_tgi <- x1 %>% filter(stimuli == 'tgi')
    # finally, replace a proportion of the data with zeros! 
    x_cnt$rating[sample(nrow(x_cnt),zp_cnt)] <- 0
    x_tgi$rating[sample(nrow(x_tgi),zp_tgi)] <- 0

    #compile both data-sets again
    x2 <- rbind(x_cnt, x_tgi)

  } else if (vas[1] == 'warm'){
    x1 <- x %>%  filter(quality == 'warm')
    # add this number to the proportion of zeros calculated for each quality & stimuli from pilot data, for each       participant
    zp_cnt <- CNTw_zero$prop[2] + r[1] 
    zp_tgi <- TGIw_zero$prop[2] + r[2]
    # adding if loop to make sure zp does not reach 0
    if (isTRUE(zp_cnt < 0)){
      zp_cnt = 0
    }
    if (isTRUE(zp_tgi < 0)){
      zp_tgi = 0
    }
    # turn zp into the proportion of total columns
    zp_cnt <- round(zp_cnt*nrow(filter(x1, stimuli == 'non-tgi')))
    zp_tgi <- round(zp_tgi*nrow(filter(x1, stimuli == 'tgi')))
    # isolate TGI and non-TGI
    x_cnt <- x1 %>% filter(stimuli == 'non-tgi')
    x_tgi <- x1 %>% filter(stimuli == 'tgi')
    # finally, replace a proportion of the data with zeros! 
    x_cnt$rating[sample(nrow(x_cnt),zp_cnt)] <- 0
    x_tgi$rating[sample(nrow(x_tgi),zp_tgi)] <- 0

    #compile both data-sets again
    x2 <- rbind(x_cnt, x_tgi)

  } else if (vas[1] == 'burn') {
    x1 <- x %>%  filter(quality == 'burn')
    # add this number to the proportion of zeros calculated for each quality & stimuli from pilot data, for each participant
    zp_cnt <- CNTb_zero$prop[2] + r[1] 
    zp_tgi <- TGIb_zero$prop[2] + r[2]
    # adding if loop to make sure zp does not reach 0
    if (isTRUE(zp_cnt < 0)){
      zp_cnt = 0
    }
    if (isTRUE(zp_tgi < 0)){
      zp_tgi = 0
    }
    # turn zp into the proportion of total columns
    zp_cnt <- round(zp_cnt*nrow(filter(x1, stimuli == 'non-tgi')))
    zp_tgi <- round(zp_tgi*nrow(filter(x1, stimuli == 'tgi')))

    # isolate TGI and non-TGI
    x_cnt <- x1 %>% filter(stimuli == 'non-tgi')
    x_tgi <- x1 %>% filter(stimuli == 'tgi')
    # finally, replace a proportion of the data with zeros! 
    x_cnt$rating[sample(nrow(x_cnt),zp_cnt)] <- 0
    x_tgi$rating[sample(nrow(x_tgi),zp_tgi)] <- 0

    #compile both data-sets again
    x2 <- rbind(x_cnt, x_tgi)
  }
  return(x2)
}
```

@@

summary(mc.og)
summary(mw.og)
summary(mb.og)
```
# Simulation of data function
```{r}
simulate_subj <- function(subid) {
# trial info
trialN = 12
conditionN = 8
ratingN = 3
# creating simulation data-frame
df <- data.frame(matrix(ncol = 0, nrow = trialN*conditionN*ratingN))
df$id <- rep(subid,trialN*conditionN*ratingN)
df$trialn <- rep(1:trialN,conditionN*ratingN)
df$dermatome <- rep(rep(c(rep("within",trialN*2), rep("across",trialN*2)),2),ratingN)
df$cold_location <- rep(rep(c(rep("dist_rostr",trialN), rep("prox_caud", trialN)),conditionN/2),ratingN)
df$stimuli <- rep(c(rep("tgi",trialN*conditionN/2), rep("non-tgi",trialN*conditionN/2)),ratingN)
df$quality <- c(rep("cold",trialN*conditionN), rep("warm",trialN*conditionN), rep("burn",trialN*conditionN))

@@

# First, simulate one data-set and run models (warm, cold and burn perception) to take a look at data
```{r}
df_sim <- data.frame()
n = 50 #an example number of participants
for (i in 1:n) {
  df <- simulate_subj(i)
  # simulating proportion of zero responses into the data using prop_zero (defined above)
  # this is a bit raw, but works!
  df_c <- prop_zero(df, 'cold')
  df_w <- prop_zero(df, 'warm')
  df_b <- prop_zero(df, 'burn')
  # bind all
  df_z <- rbind(df_c, df_w)
  df_z <- rbind(df_z, df_b)

  df_sim <- rbind(df_sim, df_z)

  #rm(df_c, df_w, df_b, df_z, df) #making sure these data-frames don't overlap
}
df50 <- df_sim
# sanity check that all simulations are different
sc <- aggregate(rating~id*stimuli, mean, data = df50)

# Plotting data distributions to see what kind of model family is required
df50$quality <- factor(df50$quality, 
                             levels = c("cold", "warm", "burn"))
ggplot2::ggplot(data = df50) +
  geom_histogram(aes(rating, fill = stimuli), alpha = .5, binwidth = .05) +
  facet_wrap(~quality) +
  theme_classic()
```

@@

# Look at model assumptions
```{r}
# cold assumption - on n = 50
mc.assump <- simulateResiduals(mc, n = 1000)
plot(mc.assump)
# warm assumption - on n = 50
mw.assump <- simulateResiduals(mw, n = 1000)
plot(mw.assump)
# warm assumption - on n = 50
mb.assump <- simulateResiduals(mb, n = 1000)
plot(mb.assump)
```

## SIMULATIONS ##
# Run through n = 1000 simulations of the pilot data, and extract p-value for the smallest effect (cold_location) of each model
# Extract the power from each model - first cold
```{r}
# the warm model has the smallest effect size, so should extract power calculation for both the cold and warm only
# do this for 40 participants, for 1000 simulations
nSims <- 1000
n <- 32 # number of participants to simulate
po_c <- data.frame()
po_w <- data.frame()
for (i in 1:nSims){
  # creating individual data-frames from simulated data
  df_sim <- data.frame()
  for (i in 1:n) {
    df <- simulate_subj(i)
    # simulating proportion of zero responses into the data using prop_zero (defined above)
    # this is a bit raw, but works!
    df_c <- prop_zero(df, 'cold')
    df_w <- prop_zero(df, 'warm')
    df_b <- prop_zero(df, 'burn')
    # bind all
    df_z <- rbind(df_c, df_w)
    df_z <- rbind(df_z, df_b)

    df_sim <- rbind(df_sim, df_z)
  }
  # then run the cold model based on simulated data
  m <- model(df_sim, 'cold')
  m1 <- summary(m)
  # extracting the p-value and effect size of cold location (our effect of interest)
  tmp <- c(m1$coefficients$cond[4,4], m1$coefficients$cond[4,1])
  po_c <- rbind(po_c,tmp)

  # and the warm model based on simulated data
  m <- model(df_sim, 'warm')
  m2 <- summary(m)
  # extracting the p-value and effect size
  tmp1 <- c(m2$coefficients$cond[4,4], m2$coefficients$cond[4,1])
  po_w <- rbind(po_w,tmp1)
}

# first get cold power
# rename columns
names(po_c)[1] <- 'pval'
names(po_c)[2] <- 'eff'
po_c$sig <- as.numeric(po_c$pval < .05) #is significant
po_c <- po_c[!is.na(po_c$sig) ,] #removing any NAs (models that did not converge) from the data frame

power_c <- sum(po_c$sig)/length(po_c$sig) # calculating power

write.csv(po_c, 'simulation_model_cold.csv', row.names = FALSE)
print(paste0('Cold power: ', round(power_c,2)))

# then for warm
# some renaming
names(po_w)[1] <- 'pval'
names(po_w)[2] <- 'eff'
po_w$sig <- as.numeric(po_w$pval < .05) #is significant
po_w <- po_w[!is.na(po_w$sig) ,] #removing any NAs (models that did not converge) from the data frame

power_w <- sum(po_w$sig)/length(po_w$sig) # calculating power

write.csv(po_w, 'simulation_model_warm.csv', row.names = FALSE)
print(paste0('Warm power: ', round(power_w,2)))
```
